Forecasting Gamma-Ray Bursts using Gravitational Waves

### 2018-11-21 Paper/Manuscript.tex
Minor/final. It is ready for referees modulo prettying up figures and typos

The science goals of the 3G detectors is very exciting. For example,
Einstein Telescope will be able to (i) distinguish between different neutron star equations of state, (ii) conduct high-precision measurements 
of source parameters to test alternate theories of gravity,
%(projected SNR of GW150914 at ET is 800, at CE 2400 vs. ALV 25)
(iii) detect binary black hole populations out to redhifts of $\sim 15$
hence  (a) measure the Hubble parameter, dark matter and dark energy densities, as well as the dark energy equation-of-state parameter;
(b) study the cosmological evolution of the stellar mass black holes (see Sec.~2 of Ref.~\cite{ET_doc} for an excellent summary and further details).

The prospect of having extremely sensitive 3G GW detectors begs a simple question: can we detect the inspirals of binary neutron star systems
early enough to witness gamma-ray bursts as they happen? In other words, can we forecast GRBs using GW detections?
The short answer is ``yes''. The long answer is still ``yes'', but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe at best $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.

@@

Let us conclude this subsection with a brief discussion on 
%our particular choice to represent the sensitivities of the near-future ALV network.
an alternate approach to construct a model for the near-future ALV network.
This involves selecting one IFO as the ``base'' detector, e.g., LIGO Livingstone then running Monte-Carlo simulations over the four random angles $\{\theta,\phi,\psi,\iota\}$. 
We would subsequently compute the quality factor of the base IFO using the
simulation results for the four angles.
We can then obtain the quality factors of the remaining two IFOs 
--- in this particular case, LIGO Hanford and Virgo --- via their positions and orientations with respect to the base IFO. Given that the four angles
are completely random, there should be no preferred point in the angle parameter space, thus the result of the Monte-Carlo runs should more or less agree with the RMS average. Correspondingly, we would expect one of the other two IFOs to have somewhat higher sensitivity, and the other one to have lower sensitivity.
This network can be further expanded using exact positions of LIGO India
and KAGRA detectors. The expansion of the detector network from three to five IFOs will offer significant improvements in source localization.
However, given that the total network SNR is a sum of quadratures, the 
expanded network will not accumulate significantly higher SNR than
our model ALV network. As such our model network should serve as a good proxy for the advance warning time computations.

@@

the fastest observed NS spin in a BNS system is $\lesssim 0.04-0.05$ \cite{0264-9381-26-7-073001, Stovall:2018ouw}.

@@

In mid 2020s LIGO will first be upgraded to A+ then to Voyager. 
However, both versions will have comparable sensitivities to Advanced LIGO
at low frequencies (see Fig.~1 of Ref.~\cite{Mills:2017urp}).
Therefore, they will not start accumulating significant SNR until $f\approx 10\,$Hz, thus offering us gains of mere minutes compared to $T_\text{AW}$ in Table~\ref{table:LIGO2020}. For this reason, we move on to 2030s: the 
era of Einstein Telescope.

@@

\begin{table}[h]
\centering
\begin{tabular}{lccccccc}
%toprule
\hline\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{7mm} $T_\text{AW}$ \ \hspace{7mm} & $\bar{\rho}_{F}$ &{} & $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{5mm} $T_\text{AW}\hspace{12mm}$& $\bar{\rho}_{F}$\T\B\\
%midrule
100 & $\approx\,$6.72 &  47.0\,minutes & 306 &{\qquad} & $\approx\,$3.27 & 5.34\,hours\ & 365\T\\
200 & $\approx\,$11.2 & 11.6\,minutes & 152 &{\qquad} & $\approx\,$4.10 & 2.87\,hours\ & 182 \\
400 & $\approx\,$18.2 & 3.00\,minutes & 75.7 &{\qquad} & $\approx\,$5.06 & 1.51\,hours\ & 90.5\\
1000 & $\approx\,$41.3 &17.2\,seconds & 29.8& \qquad &    $\approx\,$6.76 & 35.6\,minutes & 35.6  \\
% OLD version
%100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
%200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
%400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
%1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope. 
%We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer to the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}_\text{ET}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime compared to ET-B as is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\gtrsim 10\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to \emph{five hours} of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's range) will persist in the ET-C sensitivity band for a few hours after their initial detection ($T_\text{AW} > 2\,$ hours).
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET-C. 
Our values for $T_\text{AW}$ for 200 and 400\,Mpc are roughly consistent
with the findings of Ref.~\cite{Chan:2018csa}.

@@

inside a radius of $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ \cite{Mills:2017urp}, but mostly after the merger.

@@

On the other hand, ET with CE together will localize 

@@

within 200, 400\,Mpc to $1\,$deg$^2$, respectively \cite{Chan:2018csa},

@@

since CE will be sensitive enough to start accumulating SNR at $f\approx 5\,$Hz from BNSs within 400\,Mpc corresponding to $T_\text{AW}\gtrsim 1.5\,$hours.
Even without the help of a second detector, it will be possible

@@

the Earth's rotation because these nearby sources will spend hours in the ET-C band \cite{Zhao:2017cbb}.

@@

electromagnetic observations of at least dozens of gamma-ray bursts and the subsequent kilonovae
\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational in mid 2030s,

@@

%\section{Binary black hole inspirals}\label{sec:BH_BH}
For the curious reader, we briefly repeat our computations for binary black hole (BBH) inspirals.
As heavier systems inspiral too fast, we limit ourselves
to $5M_\odot-5M_\odot$ BBHs and only consider only ET-C.
For these systems, we find that $T_\text{AW} \lesssim 1\,$hour for $D\gtrsim 190\,$Mpc, but $\bar{\rho}_F$ exceeds 500 due to the increased total mass. Even for a source at 100\,Mpc, we obtain $T_\text{AW}\approx 35\,$minutes at best.
As such lightweight black holes
are expected to be rare and given the current BBH merger rate of 12-213\,Gpc$^{-3}$yr$^{-1}$ \cite{Abbott:2017vtc}, which translates to $\ord(0.1) - \ord(1)$ BBH mergers yearly
within 190\,Mpc, we statistically expect no forecasts of these events. 
Nonetheless, ET and CE will detect hundreds to thousands of BBH mergers out to redshifts of $z> 10$.

>>>>

The science goals of the 3G detectors is very exciting. For example,
Einstein Telescope will be able to (i) distinguish between different neutron star equations of state, (ii) conduct high-precision measurements 
of source parameters to test alternate theories of gravity,
%(projected SNR of GW150914 at ET is 800, at CE 2400 vs. ALV 25)
(iii) detect binary black hole populations out to redhifts of $\sim 15$
hence  (a) measure the Hubble parameter, dark matter and dark energy densities, as well as the dark energy equation-of-state parameter;
(b) study the cosmological evolution of the stellar mass black holes (see Sec.~2 of Ref.~\cite{ET_doc} for an excellent summary of the science goals and further details).

The prospect of having extremely sensitive 3G GW detectors begs a simple question: can we detect the inspirals of binary neutron star systems
early enough to witness gamma-ray bursts and the birth of the subsequent kilonovae as they happen? 
In other words, can we forecast GRBs using GW detections?
The short answer is ``yes''. The long answer is still ``yes'', but depends on many factors such as source distance and sky position, detector sensitivity and orientation. One also needs to account for the fact that GRBs are thought to be collimated emissions with narrow outburst angles \cite{Kumar:2014upa}.
As such, most GRBs that occur in the universe are electromagnetically undetectable. Fortunately, this was not the case with GRB170817A, 
but even for a reasonable GRB jet opening angle of $\sim 10^\circ$, we expect to observe at best $\sim 10\%$ of neutron star mergers as GRBs \cite{Patricelli:2016bkt}. 
However, it is possible that some of the near-Earth pointing short hard GRBs may be observed as long-duration GRBs or X-ray flashes thus increasing this percentage \cite{Bucciantini:2011kx}.
On the other hand, the optical emission from the kilonova should be less collimated, thus much more likely
to be detected \cite{Troja:2017nqp}.

@@

Let us conclude this subsection with a brief discussion on 
%our particular choice to represent the sensitivities of the near-future ALV network.
an alternate approach to construct a model for the near-future ALV network.
This involves selecting one IFO as the ``base'' detector, e.g., LIGO Livingstone, then running Monte-Carlo simulations over the four random angles $\{\theta,\phi,\psi,\iota\}$. 
We would subsequently compute the quality factor of the base IFO using the
simulation results for the four angles.
We can then obtain the quality factors of the remaining two IFOs 
--- in this particular case, LIGO Hanford and Virgo --- via their positions and orientations with respect to the base IFO. Given that the four angles
are completely random, there should be no preferred region in the angle parameter space, thus the result of the Monte-Carlo runs should more or less agree with the RMS average. Correspondingly, we would expect one of the other two IFOs to have somewhat higher sensitivity, and the other one to have lower sensitivity.
This network can be further expanded using exact positions of LIGO India
and KAGRA detectors. The expansion of the detector network from three to five IFOs will offer significant improvements in source localization.
However, given that the total network SNR is a sum of quadratures, the 
expanded network will not accumulate significantly higher SNR than
our model ALV network. As such our model network should serve as a good proxy for the advance warning time computations.

@@

the fastest observed NS spin in a BNS system is $\approx 0.04-0.05$ \cite{0264-9381-26-7-073001, Stovall:2018ouw}.

@@

In mid 2020s LIGO will first be upgraded to A+ then to Voyager. 
However, both versions will have comparable sensitivities to Advanced LIGO
at low frequencies (see Fig.~1 of Ref.~\cite{Mills:2017urp}).
Therefore, they will not start accumulating significant SNR until $f\approx 10\,$Hz, thus offering us gains of a few minutes over $T_\text{AW}$ in Table~\ref{table:LIGO2020}. For this reason, we move on to 2030s: the 
era of Einstein Telescope.

@@

\begin{table}[h]
\centering
\begin{tabular}{lccccccc}
%toprule
\hline\hline
$D\,$(Mpc) & \multicolumn{3}{c}{ET-B} &  & \multicolumn{3}{c}{ET-C}\T\B\\
\hline
{}& $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{7mm} $T_\text{AW}$ \ \hspace{7mm} & $\bar{\rho}_{F}$ &{} & $\bar{f}_\text{ET}\,$(Hz) & \ \hspace{5mm} $T_\text{AW}\hspace{12mm}$& $\bar{\rho}_{F}$\T\B\\
%midrule
100 & $\approx\,$6.72 &  47.0\,minutes & 306 &{\qquad} & $\approx\,$3.27 & 5.34\,hours\ & 365\T\\
200 & $\approx\,$11.2 & 11.6\,minutes & 152 &{\qquad} & $\approx\,$4.10 & 2.87\,hours\ & 182 \\
400 & $\approx\,$18.2 & 3.00\,minutes & 75.7 &{\qquad} & $\approx\,$5.06 & 1.51\,hours\ & 90.5\\
1000 & $\approx\,$41.3 &17.2\,seconds & 29.8& \qquad &    $\approx\,$6.76 & 35.6\,minutes & 35.6  \\
% OLD version
%100 & 6.7237 &  2818 & 306.0 &{\qquad} & 3.2692 & 19212 & 365.4\T\\
%200 & 11.2104 & 698.6 & 152.4 &{\qquad} & 4.1028 & 10138 & 182.1 \\
%400 & 18.2234 & 179.8 & 75.74 &{\qquad} & 5.0607 & 5437 & 90.48\\
%1000 & 41.3537 &17.2 & 29.79& \qquad & 6.7553 & 2138 & 35.60  \\
\hline\hline
%bottomrule
\end{tabular}
\caption{Forecasting capabilities of Einstein Telescope. 
%We can expect over \emph{five} hours of early warning for nearby sources. 
ET-B and ET-C refer to the different configurations shown in Fig.~\ref{fig:ETB2030}. For the advance warning times, we only present the result of the more accurate 3.5PN computation. $\bar{f}_\text{ET}$ is the threshold frequency at which
ET-B/C accumulate SNR of 15 which we take to be our detection criterion. Note that both $T_\text{AW}$ and $\bar\rho_F$ are larger for ET-C
due to its improved sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime compared to ET-B as is clear in Fig.~\ref{fig:ETB2030}.
These results and those of Table.~\ref{table:LIGO2020} are summarized in Fig.~\ref{fig:summary}.}\label{table:ET}
\end{table}
%
%
As the results of Table~\ref{table:ET} indicate, ET-B will provide $\gtrsim 10\,$ minutes of warning time for BNS inspirals at $D\lesssim 200\,$Mpc.
Instead of speculating whether or not this is enough to forecast GRBs, we turn our attention to ET-C's forecasting capabilities.
As can be seen from Fig.~\ref{fig:ETB2030}, ET-C's increased sensitivity in the $1\,\text{Hz}\lesssim f\lesssim 30\,$Hz regime will significantly lengthen the advance warning times.
This is best highlighted by the sixth colum of Table~\ref{table:ET}, 
where we show that ET-C will provide up to \emph{five hours} of early warning time for nearby ($D\lesssim 100\,$Mpc) neutron star mergers.
Moreover, BNS inspirals as far as 200\,Mpc (the ALV network's range) will persist in the ET-C sensitivity band for a few hours after their initial detection ($T_\text{AW} > 2\,$ hours).
Current BNS merger rate, inferred from A-LIGO's O1-O2 periods, implies that $\sim 40$ to $\sim600$ sources within a volume of $(200\,\text{Mpc})^3$ will be detected yearly by ET-C. 
Our values for $T_\text{AW}$ for $D=200, 400\,$Mpc are roughly consistent
with the findings of Ref.~\cite{Chan:2018csa}.

@@

inside a radius of $D\approx 1000\,$Mpc to within $ 10\,\text{deg}^2$ \cite{Mills:2017urp}, but mostly \emph{after} the merger.

@@

On the other hand, ET together with CE will localize 

@@

within 200, 400\,Mpc to $1\,$deg$^2$, respectively \cite{Chan:2018csa}.

@@

This is thanks to CE's low-frequency sensitivity which is sufficient to start accumulating SNR at $f\approx 5\,$Hz from BNSs within 400\,Mpc corresponding to $T_\text{AW}\gtrsim 1.5\,$hours.
Even without the help of a second detector, ET-C will be able

@@

the Earth's rotation because these nearby sources will spend hours in its frequency band \cite{Zhao:2017cbb}.

@@

electromagnetic observations of at least dozens of gamma-ray bursts and the subsequent kilonovae\footnote{As the ET construction/commissioning timeline is currently uncertain, we can only speculate that ET-C will be operational in mid 2030s,

@@

%\section{Binary black hole inspirals}\label{sec:BH_BH}
For the curious reader, we briefly repeat our computations for binary black hole (BBH) inspirals.
As heavier systems inspiral too fast, we limit ourselves
to $5M_\odot-5M_\odot$ BBHs and only consider ET-C.
For these systems, we find that $T_\text{AW} \gtrsim 1\,$hour only for $D\lesssim 190\,$Mpc, with $\bar{\rho}_F$ exceeding 500 due to the increased total mass. Even for a source at 100\,Mpc, we obtain $T_\text{AW}\approx 35\,$minutes at best.
As such lightweight black holes
are expected to be rare and given the current BBH merger rate of 12-213\,Gpc$^{-3}$yr$^{-1}$ \cite{Abbott:2017vtc}, which translates to $\ord(0.1) - \ord(1)$ BBH mergers yearly
within 190\,Mpc, we statistically expect no forecasts of these events. 
Nonetheless, ET and CE will detect hundreds to thousands of BBH mergers out to redshifts of $z> 10$.
